Colors and Residual-based Shadings in the Strucplot Framework

### 2005-10-18 trunk/vcd/inst/doc/shading.Rnw
Misc KH changes.
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/vcd@475 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

\begin{enumerate}
\item Pre-computing the graphical parameters (e.g., fill colors),
  encapsulating them into an object of class
  \class{gpar} as demonstrated in the previous section, and
  passing this object to the \code{gp} argument.
\item Providing a grapcon function to the \code{gp} argument that 
  takes residuals as input and returns an object as described in
  alternative 1.
\item Providing a grapcon \emph{generating} function taking parameters and returning
  a function as described in alternaive 2. 
\end{enumerate}

@@

We simply wrap our grapcon shading function in another 
function that takes all additional arguments it needs to use, possibly
preprocesses them, and returns the actual shading function. This returned
function will have access to the parameters since in {\sf R}, nested
functions are lexically scoped. Thus, the grapcon generator
returns (`creates') a `parameterized' shading function with the minimal standard interface
\codefun{strucplot} requires. The follwing example shows the necessary
extensions for our running example:

@@

\noindent In the call of \codefun{mosaic} (figure not shown), 
using the new \codefun{shading3a\_fun} function, we can now simply change the colors:

@@

\noindent The procedure described so far is a rather general concept,
applicable to a wide family of user-level \pkg{grid} graphics. Indeed, the
customization of other components of the strucplot framework
(labeling, spacing, legend, and core functions) follows the same idea.
Now for the shading functions, more customization is needed. Note that
\codefun{shading3a\_fun} needs to be evaluated by the user, even if the
defaults shall be used. It is a better idea to let \codefun{strucplot} call
the generating function, which, in particular, allows the passing of arguments that
are computed by \codefun{strucplot}. Since shading
functions can be used for visualizing significance (see Section 
\ref{sec:overview}), it makes sense for generating functions to have access to the model,
i.e., observed and expected values, residuals, and degrees of
freedom. For example, the \codefun{shading\_max} generating function
computes a permutation distribution of the maximum statistic and $p$
values for specified significance levels 
based on the observed table to create data-driven cut-off
points. If this was done in the shading function itself, the
permutation statistic would be recomputed every time the shading
function is called, resulting in possibly severe performance loss 
and numerical inconsistencies. Therefore, generating functions for
shadings are required to take at least the parameters 
\code{observed}, \code{expected}, \code{residuals}, and \code{df}
(these are provided by the strucplot framework),
followed by other parameters controlling the shading
appearance (to be specified by the user):

@@

In some sence, generating
functions for shadings are parameterized both by the user and the
strucplot framework. For shading functions that require model
information, the user-specified parameters are 
to be passed to the \code{gp\_args} argument instead, and for this to
work, the generating function needs a class attribute to be distinguishable from
the ``normal'' shading functions. For others
(like our simple \codefun{shading3b\_fun}) this is optional, but
recommended for consistency:

>>>>

\begin{enumerate}
\item Precomputing the graphical parameters (e.g., fill colors),
  encapsulating them into an object of class
  \class{gpar} as demonstrated in the previous section, and
  passing this object to the \code{gp} argument.
\item Providing a grapcon function to the \code{gp} argument that 
  takes residuals as input and returns an object as described in
  alternative 1.
 \item Providing a grapcon \emph{generating} function taking parameters
  and returning a function as described in alternative~2.
\end{enumerate}

@@

We simply wrap our grapcon shading function in another 
function that takes all additional arguments it needs to use, possibly
preprocesses them, and returns the actual shading function. This returned
function will have access to the parameters since in {\sf R}, nested
functions are lexically scoped. Thus, the grapcon generator
returns (`creates') a `parameterized' shading function with the minimal standard interface
\codefun{strucplot} requires. The following example shows the necessary
extensions for our running example:

@@

\noindent In the call to \codefun{mosaic}, using the new
\codefun{shading3a\_fun} function, we can now simply change the colors:

@@

\noindent (figure not shown).  The procedure described so far is a
rather general concept, applicable to a wide family of user-level
\pkg{grid} graphics. Indeed, the customization of other components of
the strucplot framework (labeling, spacing, legend, and core functions)
follows the same idea.  Now for the shading functions, more
customization is needed. Note that \codefun{shading3a\_fun} needs to be
evaluated by the user, even if the defaults are to be used. It is a
better idea to let \codefun{strucplot} call the generating function,
which, in particular, allows the passing of arguments that are computed
by \codefun{strucplot}. Since shading functions can be used for
visualizing significance (see Section \ref{sec:overview}), it makes
sense for generating functions to have access to the model, i.e.,
observed and expected values, residuals, and degrees of freedom. For
example, the \codefun{shading\_max} generating function computes a
permutation distribution of the maximum statistic and $p$ values for
specified significance levels based on the observed table to create
data-driven cut-off points. If this was done in the shading function
itself, the permutation statistic would be recomputed every time the
shading function is called, resulting in possibly severe performance
loss and numerical inconsistencies. Therefore, generating functions for
shadings are required to take at least the parameters \code{observed},
\code{expected}, \code{residuals}, and \code{df} (these are provided by
the strucplot framework), followed by other parameters controlling the
shading appearance (to be specified by the user):

@@

In some sense, generating
functions for shadings are parameterized both by the user and the
strucplot framework. For shading functions that require model
information, the user-specified parameters are 
to be passed to the \code{gp\_args} argument instead, and for this to
work, the generating function needs a class attribute to be distinguishable from
the ``normal'' shading functions. For others
(like our simple \codefun{shading3b\_fun}) this is optional, but
recommended for consistency:
